---
title: "Blog Post 3"
author: "Group 5" 
date: "2024-10-28"
date-modified: "2024-10-28"
draft: FALSE
---
```{r}
if (!dir.exists("../ma-4615-fa24-final-project-group-5/dataset")) {
  dir.create("../ma-4615-fa24-final-project-group-5/dataset")
}

csv_employment_data <- read.csv("../ma-4615-fa24-final-project-group-5/dataset/Employment2017.csv")

saveRDS(csv_employment_data, "../ma-4615-fa24-final-project-group-5/dataset/Employment2017.rds")

employment_data <- readRDS("../ma-4615-fa24-final-project-group-5/dataset/Employment2017.rds")
```

```{r}
#summary(employment_data)
sum(is.na(employment_data))
```
There are 25763 missing values. Many asterisks are used to protect individuals (employee or establishment) from the risk of identification. An asterisk is sometimes used to protect a vulnerable cell and other times to safeguard another at-risk cell.



Next, we create 3 charts to help with data visualization. 
The first bar chart helps us visualize the 10 different job types labeled in the dataset by summing up all the rows within the given columns starting with "TOTAL_" and finding their percentages.
The second bar chart does something similar, but this time we want to see the proportionality of the 7 given race variables disregarding job types. In a future blog post, we will look into their proportionality with the different job types. 
The third pie chart looks at the gender variation (male/female0 between all jobs and races. 

```{r p1}
library(tidyverse)

employment_data <- readRDS("Employment2017.rds")
employment_data[employment_data == "*"] <- NA

total_columns <- c("TOTAL1", "TOTAL2", "TOTAL3", "TOTAL4", "TOTAL5", "TOTAL6", "TOTAL7", "TOTAL8", "TOTAL9", "TOTAL1_2")

employment_data[total_columns] <- lapply(employment_data[total_columns], function(x) as.numeric(x))

employment_data[total_columns] <- 
  lapply(employment_data[total_columns], function(x) replace_na(x, 0))

total_counts <- colSums(employment_data[total_columns], na.rm = TRUE)

proportion_df <- data.frame(Job_Type = names(total_counts), Count = total_counts)

proportion_df$Job_Type <- recode(proportion_df$Job_Type,
  TOTAL1 = "Senior Officers and Managers",
  TOTAL2 = "Professionals",
  TOTAL3 = "Technicians",
  TOTAL4 = "Sales Workers",
  TOTAL5 = "Clericals",
  TOTAL6 = "Craft",
  TOTAL7 = "Operatives",
  TOTAL8 = "Laborers",
  TOTAL9 = "Services",
  TOTAL1_2 = "Mid Officers and Managers"
)

proportion_df <- proportion_df %>%
  mutate(Proportion = Count / sum(Count))

ggplot(proportion_df, aes(x = Job_Type, y = Proportion, fill = Job_Type)) +
  geom_bar(stat = "identity") +  
  labs(title = "Proportionality of Jobs by Type", 
       x = "Job Type", 
       y = "Proportion") +
  theme_gray() +  
  scale_y_continuous(labels = scales::percent) +  
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) 
```
```{r p2}
race_columns <- c("WHT10", "BLKT10", "HISPT10", "ASIANT10", "AIANT10", "NHOPIT10", "TOMRT10")

employment_data[race_columns] <- lapply(employment_data[race_columns], function(x) as.numeric(x))

employment_data[race_columns] <- 
  lapply(employment_data[race_columns], function(x) replace_na(x, 0))

total_race_counts <- colSums(employment_data[race_columns], na.rm = TRUE)

race_proportion_df <- data.frame(Race = names(total_race_counts), Count = total_race_counts)

race_proportion_df$Race <- recode(race_proportion_df$Race,
  WHT10 = "White",
  BLKT10 = "Black or African American",
  HISPT10 = "Hispanic",
  ASIANT10 = "Asian",
  AIANT10 = "American Indian or Alaska Native",
  NHOPIT10 = "Native Hawaiian or Other Pacific Islander",
  TOMRT10 = "Two or more Races"
)

race_proportion_df <- race_proportion_df %>%
  mutate(Proportion = Count / sum(Count))

ggplot(race_proportion_df, aes(x = Race, y = Proportion, fill = Race)) +
  geom_bar(stat = "identity") +  
  labs(title = "Proportionality of Race", 
       x = "Race", 
       y = "Proportion") +
  theme_gray() +  
  scale_y_continuous(labels = scales::percent) +  
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) 
```

```{r p3}
gender_columns <- c("MT10", "FT10")

employment_data[gender_columns] <- lapply(employment_data[gender_columns], function(x) as.numeric(x))

employment_data[gender_columns] <- 
  lapply(employment_data[gender_columns], function(x) replace_na(x, 0))

total_gender_counts <- colSums(employment_data[gender_columns], na.rm = TRUE)

gender_proportion_df <- data.frame(Gender = names(total_gender_counts), Count = total_gender_counts)

gender_proportion_df$Gender <- recode(gender_proportion_df$Gender,
  MT10 = "Male", FT10 = "Female")

gender_proportion_df <- gender_proportion_df %>%
  mutate(Proportion = Count / sum(Count),
         Percentage = Proportion * 100)  

ggplot(gender_proportion_df, aes(x = "", y = Proportion, fill = Gender)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") + 
  labs(title = "Proportionality of Gender", 
       fill = "Gender") +
  theme_gray() +  
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank()) +
  geom_text(aes(label = paste0(round(Percentage, 1), "%")), 
            position = position_stack(vjust = 0.5)) 
```
Next, we create our cleaned folder. It is also included in the data.qmd as well as the scripts/cleaned_data file. 
```{r p4}
library(dplyr)
library(tidyselect)

employment_data <- readRDS("Employment2017.rds")

cleaned_data <- employment_data %>%
  select(
    Nation, Region, Division, State, CBSA, County, NAICS2, NAICS2_Name, NAICS3, NAICS3_Name, Establishments,
    matches("^WHT_([1-9]|1_2)$"),
    matches("^BLKT_([1-9]|1_2)$"),
    matches("^HISPT_([1-9]|1_2)$"),
    matches("^ASIANT_([1-9]|1_2)$"),
    matches("^AIANT_([1-9]|1_2)$"),
    matches("^NHOPIT_([1-9]|1_2)$"),
    matches("^TOMRT_([1-9]|1_2)$"),
    matches("^MT_([1-9]|1_2)$"),
    matches("^FT_([1-9]|1_2)$")
  )

if (!dir.exists("../ma-4615-fa24-final-project-group-5/dataset")) {
  dir.create("../ma-4615-fa24-final-project-group-5/dataset", recursive = TRUE)
}

saveRDS(cleaned_data, "../ma-4615-fa24-final-project-group-5/dataset/Cleaned_Employment2017.rds")
```

Data for Equity

Transparency
Transparency is a critical component in equitable data analysis. In our project, being transparent would provide us with any potential limitations in the dataset, such as the lack of representation for smaller sized companies. This exclusion could skew our analysis, if smaller sized companies employ more people from certain demographic groups or adhere to different hiring practices. 

Privacy and Data Protection
In our dataset, the use of asterisks (*) in cells are designed to prevent individual identification, which aligns with the Privacy and Data Protection principle by protecting vulnerable populations. Doing so avoids any attempts to infer or disaggregate data that could lead to re-identifying individuals or small groups, particularly in fields where there is limited representation (e.g., certain job categories or minority groups). This respect for privacy aligns with data equity by fostering a safe environment for marginalized communities.

Inclusivity and Representativeness
In our dataset, focusing on medium sized companies disregards the smaller sized companies, which could include potentially significance in minority or gender-represented employment. It is important to include all data as it better captures a representation of the dataset for all demographic groups. By including data on smaller sized companies, we are able to broaden the analysis and provide a more complete picture of industry disparities and representation across all workforce segments. This is particularly significant as many family businesses, which fall beyond the purview of our dataset, tend to be minority owned.